---
title: "How is No Crowd Effecting the English Premier League?"
subtitle: "A unique insight into footballs 12th man"
author: "Matthew Henderson"
date: "20 February 2021"
output: ioslides_presentation
df_print: "kable"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Motivation

- The 2020/21 English Premier League season is the first to be played with little to no crowd in any matches.
- Unique opportunity to analyse whether having a crowd effects a football match.
- Great English Premier League dataset contanaing seasons 1993/94 - 2020/21 on kaggle <https://www.kaggle.com/irkaal/english-premier-league-results>
- Seasons 1993/94 and 1994/95 were dropped since there was 22 teams opposed to 20 after.

## Summary Stats 

```{r, echo=FALSE, message=FALSE}
## How has no crowd effected the Premier League this Season?
library(dplyr)
library(moments)
library(tibble)
# read in results data from Seasons 1993/94 - now
results <- read.csv("data/external/results/results.csv")
results <- results %>% filter(Season != "1993-94")
results <- results %>% filter(Season != "1994-95")
results <- results %>% select(Season, DateTime, HomeTeam, AwayTeam, FTHG, FTAG, FTR, Referee)
# Create statistical summaries from this season
HomeSummary <- results %>% group_by(Season) %>% summarise(mean = mean(FTHG), median = median(FTHG), sd = sd(FTHG), skew = skewness(FTHG), min = range(FTHG)[1], max = range(FTHG)[2])
AwaySummary <- results %>% group_by(Season) %>% summarise(mean = mean(FTAG), median = median(FTAG), sd = sd(FTAG), skew = skewness(FTAG), min = range(FTAG)[1], max = range(FTAG)[2])
HomeSummary$HoA <- "Home"
AwaySummary$HoA <- "Away"
GoalSummary <- rbind(HomeSummary, AwaySummary)
GoalSummary$HoA <- factor(GoalSummary$HoA, levels = c("Home", "Away"))
GoalSummary <- arrange(GoalSummary, Season)
GoalSummary <- mutate(GoalSummary, ID = "")
GoalSummary$ID[GoalSummary$Season != "2020-21"] <- "Past"
GoalSummary$ID[GoalSummary$Season == "2020-21"] <- "Present"
GoalSummary$ID <- factor(GoalSummary$ID, levels = c("Past", "Present"))
tail(GoalSummary, 12)
```


## Average goals scored at Home vs Away

```{r, echo=FALSE}
library(ggplot2)
# create a plot to compare the mean goals this season against previous seasons
g <- ggplot(GoalSummary, aes(x = HoA, y = mean, group = Season))
g <- g + geom_line(size = 1, aes(colour = ID)) + geom_point(size = 5, pch = 21, fill = "salmon", alpha = 0.5, data = filter(GoalSummary, ID == "Past"))
g <- g + geom_point(size = 10, pch = 21, fill = "black", data = filter(GoalSummary, ID == "Present"))
g
```

## Average goals scored at Home vs Away

- The 2020/21 Season is the only in the last 25 years to have a near horizontal trend.
- The lowest average of home goals.
- The highest average of away goals.
- Can we prove this season is unique and not a product of random chance?

## Statistical Proof

- $H_0:$ Premier League teams score the same number of goals at home as they do away from home on average.
- $H_a:$ Premier League teams score more goals at home than they do away from home on average.

Formally:

$$
H_0: \mu = 0
$$

Versus:

$$
H_0: \mu > 0
$$

## Statistical Proof

```{r, echo=FALSE}
with(filter(results, Season != "2020-21"), t.test(FTHG, FTAG, paired = TRUE, alternative = "greater"))
```

## What is happening exactly?

```{r, echo=FALSE, message=FALSE}
library(reshape2)
# Now lets look at the change in distribution
# home ... 
HomeDistr <- results %>% group_by(Season) %>% count(FTHG)
HomeDistrWide <- dcast(HomeDistr, Season ~ FTHG, value.var = "n")
HomeDistrWide[is.na(HomeDistrWide)] <- 0
HomeDistr <- melt(HomeDistrWide, id.vars = "Season", value.name = "n", variable.name = "FTHG")
HomeDistr <- HomeDistr %>% arrange(Season, FTHG)
# away ... 
AwayDistr <- results %>% group_by(Season) %>% count(FTAG)
AwayDistrWide <- dcast(AwayDistr, Season ~ FTAG, value.var = "n")
AwayDistrWide[is.na(AwayDistrWide)] <- 0
AwayDistr <- melt(AwayDistrWide, id.vars = "Season", value.name = "n", variable.name = "FTAG")
AwayDistr <- AwayDistr %>% arrange(Season, FTAG)
#
names(HomeDistr) <- c("Season", "FTG", "n"); names(AwayDistr) <- c("Season", "FTG", "n")
HomeDistr$HoA <- "Home"; AwayDistr$HoA <- "Away"
GoalDistr <- rbind(HomeDistr, AwayDistr)
GoalDistr$HoA <- factor(GoalDistr$HoA, levels = c("Home", "Away"))
# add in a probability density for each goal
GoalDistr <- GoalDistr %>% group_by(Season, HoA) %>% mutate(p = n / sum(n))
# Assuming the other seasons in the Premeir League and 2020-21 is an outlier
# lets fit a distribution to all of the other seasons and compare to 2020-21
# split home and away
GoalDistr$ID <- ""
GoalDistr$ID[GoalDistr$Season != "2020-21"] <- "Past Seasons"
GoalDistr$ID[GoalDistr$Season == "2020-21"] <- "This Season"
GoalDistr$ID <- factor(GoalDistr$ID, levels = c("Past Seasons", "This Season"))
# group by past and present and find a general distribution for the past seasons
GeneralDistr <- GoalDistr %>% group_by(ID, HoA, FTG) %>% summarise(p = mean(p)) 
GeneralDistr$FTG <- as.numeric(as.character(GeneralDistr$FTG))
# plot the distributions 
g1 <- ggplot(GeneralDistr, aes(FTG, p, fill = ID)) + 
  geom_point(size = 5, pch = 21, alpha = 0.5) + 
  geom_line(aes(linetype = ID, colour = ID), size = 1) +
  facet_wrap(~HoA, nrow = 1) +
  scale_x_continuous(breaks = 0:9)
g1
```

## Conclusions

- The average of home goals scored by Premier League teams is more than the average away goals (around 0.4 more).
- In the 2020/21 season there has been no difference between home and away goals in the  Premier League.
- Given what we know about this season we can attribute that to there being no crowd and conclude that a teams home support is in effect the 12th man in football.




